\bf Landslide: \\ Systematic Dynamic Race Detection in Kernel Space

### 2012-05-12 design.tex
continue copyediting

\begin{itemize}
	\item {\bf User-provided instrumentation.} Some parts of the kernel may be written in any number of ways, and hence require the user's assistance for Landslide to understand. The user-provided instrumentation, described in detail in Section~\ref{sec:using-instrumenting}, informs Landslide about thread-related lifecycle and scheduling events and the anatomy of the kernel's scheduler.
	\item {\bf Automatic instrumentation.} All Pebbles kernels have some things in common, and Landslide's build system is able to automatically instrument certain parts of the kernel. This instrumentation informs Landslide about the dynamic memory allocator, common library functions such as \texttt{panic}, and certain aspects of the kernel's executable format.
\end{itemize}

@@

Landslide maintains a mirror image of the guest kernel's dynamic allocation heap, so it can know at any point what memory ranges are allocated and what ranges used to be allocated but now are freed. This set is updated each time the guest kernel calls \texttt{malloc} or \texttt{free}.

@@

The arbiter identifies points during execution should count as decision points. The selection is mainly controlled by the user, during the annotation and configuration process, but it also automatically identifies {\em voluntary reschedules}, which comprise the ``minimal necessary set'' of decision points.

@@

The explorer maintains a representation of the current branch of the decision tree. It is responsible for checkpointing at each decision point, deciding at the end of the test which branch of the tree to execute next (i.e., selecting which decision point should have been decided differently), and backtracking to appropriate points in the test's execution.

@@

Landslide also makes use of Simics ``bookmarks'' to checkpoint and restore the execution state at decision points during the test.

@@

% A healthy dose of snark.
Without a formal specification of the internals of the guest kernel's implementation\cite{sel4}, it is of course impossible to identify both soundly and completely when a behaviour that constitutes a ``bug'' arises during a test case's execution.
In order to identify when the guest kernel has done something incorrect, Landslide performs several different types of checks, some accurate but noncomprehensive, and some heuristic-based.

@@

In order to establish a happens-before relation between transitions, Landslide uses the state of the scheduler runqueues (Section~\ref{sec:components-sched}), which are snapshotted at every decision point.

>>>>

\begin{itemize}
	\item {\bf User-provided instrumentation.} Some parts of the kernel may be written in any number of ways, and hence require the user's assistance for Landslide to understand. The user-provided instrumentation, described in detail in Section~\ref{sec:using-instrumenting}, informs Landslide about thread-related lifecycle and scheduling events and the anatomy of the kernel's scheduler.
	\item {\bf Automatic instrumentation.} All Pebbles kernels have some things in common (largely due to the common starter code provided for the class project), and Landslide's build system is able to automatically instrument certain parts of the kernel. This instrumentation informs Landslide about the dynamic memory allocator, common library functions such as \texttt{panic}, and certain aspects of the kernel's executable format.
\end{itemize}

@@

Landslide maintains a mirror image of the guest kernel's dynamic allocation heap, so it can know at any point which memory ranges are allocated and which ranges used to be allocated but now are freed. This set is updated each time the guest kernel calls \texttt{malloc} or \texttt{free}.

@@

The arbiter identifies points during execution that should count as decision points. The selection is mainly controlled by the user, during the annotation and configuration process. In addition, the arbiter also automatically identifies {\em voluntary reschedules}, which comprise the ``minimal necessary set'' of decision points.

@@

The explorer maintains a representation of the current branch of the decision tree. It is responsible for checkpointing the state of both Landslide and the guest kernel at each decision point, deciding at the end of the test which branch of the tree to execute next (i.e., selecting which decision point should have been decided differently), and backtracking to appropriate points in the test's execution.

@@

Landslide also makes use of Simics ``bookmarks'' to checkpoint and restore the guest kernel's execution state at decision points during the test.

@@

% A healthy dose of snark.
Without a formal specification of the internals of the guest kernel's implementation\cite{sel4}, it is impossible to identify both soundly and completely when a behaviour that constitutes a ``bug'' arises during a test case's execution.
In order to identify when the guest kernel has done something incorrect, Landslide performs several different types of checks, some accurate but noncomprehensive, and some heuristic-based.

@@

The happens-before relation expresses for each pair of transitions whether executing the first one is required to ``enable'' the second one. In order to establish the relation, Landslide uses the state of the scheduler runqueues (Section~\ref{sec:components-sched}), which are snapshotted at every decision point (Section~\ref{sec:components-explore}).

### 2012-05-12 evaluation.tex
continue copyediting

We evaluated Landslide in two ways.
First, we conducted a user study with present students of 15-410, to understand the eases and difficulties of Landslide's user interface process, to understand how to polish it into a useful debugging tool.
Second, we did a case study of six bugs in two kernels, investigating them in-depth to understand how systematic testing might best be harnessed to find bugs without in advance knowing the decision set necessary to expose them.

@@

To evaluate the experience of non-expert users working with Landslide, we met with 5 groups of students in 15-410 (9 students in total) during the final week of the kernel project, and had them use Landslide with their own kernels.
In order to garner student interest in Landslide, the author gave a lecture in 15-410 on systematic exploration in general and on Landslide specifically during the second-to-last week of the kernel project.\footnote{
The lecture slides are available at \url{http://www.cs.cmu.edu/~410-s12/lectures/L30_Landslide.pdf}, and at \url{http://bblum.net/landslide-lecture.pdf}, and at \url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-lecture.pdf}.}

@@

\begin{itemize}
	\item {\bf Special-case context switch behaviour.} One kernel had a special context-switcher for the case when a thread was vanishing, which had to be special-cased in the instrumentation.
		Some kernels had different context switch designs for just-forked threads (as discussed in the ``Thread Creation'' bullet in Section~\ref{sec:challenges-design}), and Landslide's code had to be modified during the study to support different designs.
	\item {\bf Unexpected compiler optimisations.} Listing scheduler functions in \texttt{config.landslide} could be confusing if the compiler inlined them.
	\item {\bf Driver thread interfering with test lifecycle tracking.} One kernel (Pathos, the 15-410 reference kernel) had to be modified to ensure the keyboard thread ran once before the test started, which was not mandated by the kernel design. (If the thread didn't run, it would conflict with the population-tracking described in Section~\ref{sec:components-test}, and Landslide would not be able to detect when the test was over.)
	\item {\bf Kernel behaviour requirement violations.}
		The majority of groups needed to change their scheduling around the idle thread in order to meet the requirement that the kernel never run idle when progress could otherwise be made (Sections~\ref{sec:challenges-design} and \ref{sec:using-requirements-sched}).
		Also, one kernel had to have its custom memory allocator disabled, because Landslide only supports LMM, the provided allocator.
\end{itemize}

@@

Of five student groups that we worked with during the user study, four groups completed the instrumentation process and got Landslide to explore at least a minimal decision tree. Of those groups, two found race conditions in their kernel. Additionally, Landslide helped all four groups find deterministic bugs - three groups had previously unknown problems with running the idle thread inappropriately, and the other group had a simple use-after-free.

>>>>

We evaluated Landslide in two ways.
First, we conducted a user study with students of 15-410 during the spring 2012 semester, to understand the ease and difficulty of Landslide's user interface process, to understand how to polish it into a more useful debugging tool.
Second, we did a case study of six bugs in two kernels, investigating them in-depth to understand how systematic testing might best be harnessed to find bugs without in advance knowing the decision set necessary to expose them.

@@

To evaluate the experience of non-expert users working with Landslide, we met with 5 groups of students in 15-410 (9 students in total) during the final week of the kernel project in spring 2012, and had them use Landslide with their own kernels.
In order to garner student interest in Landslide, the author gave a lecture in 15-410 on systematic exploration in general and on Landslide specifically during the second-to-last week of the kernel project.\footnote{
The lecture slides are available at \url{http://www.cs.cmu.edu/~410-s12/lectures/L30_Landslide.pdf}, and at \url{http://bblum.net/landslide-lecture.pdf}, and at \url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-lecture.pdf}.}

@@

\begin{itemize}
	\item {\bf Special-case context switch behaviour.} One kernel had a special context-switcher for the case when a thread was vanishing, which had to be special-cased in the instrumentation.
		Some kernels had different context switch designs for just-forked threads (as discussed in the ``Thread Creation'' bullet in Section~\ref{sec:challenges-design}), and Landslide's code had to be modified during the study to support different designs.
	\item {\bf Unexpected compiler optimisations.} Listing scheduler functions in \texttt{config.landslide} could be confusing if the compiler inlined them.
	\item {\bf Driver thread interfering with test lifecycle tracking.} One kernel (Pathos, the 15-410 reference kernel) had to be modified to ensure the keyboard thread ran once before the test started, which was not mandated by the kernel design. (If the thread didn't run, it would conflict with the population-tracking described in Section~\ref{sec:components-test}, and Landslide would not be able to detect when the test was over.)
	\item {\bf Kernel behaviour requirement violations.}
		The majority of groups needed to change their scheduling around the idle thread in order to meet the requirement that the kernel never run idle when progress could otherwise be made (Sections~\ref{sec:challenges-design} and \ref{sec:using-requirements-sched}).
		Also, one kernel had to have its custom memory allocator disabled, because Landslide only supports LMM, the allocator that 15-410 provides.
\end{itemize}

@@

Of five student groups that we worked with during the user study, four groups completed the instrumentation process and got Landslide to explore at least a minimal decision tree. (A minimal decision tree is one resulting from decision points only on voluntary reschedules, which Landslide automatically identifies.) Of those groups, two found race conditions in their kernel. Additionally, Landslide helped all four groups find deterministic bugs - three groups had previously unknown problems with running the idle thread inappropriately, and the other group had a simple use-after-free.

### 2012-05-12 futurework.tex
fix minor typoes
continue copyediting

\begin{figure*}[h]
	\begin{center}
	\includegraphics[width=0.75\textwidth]{wont-modify-vail.jpg}
	\end{center}
	\caption{Ex-members of Operating Systems course staff lament the difficulty involved in teaching students concurrency debugging skills.}
\end{figure*}

@@

We believe that a tool such as Landslide could change the way students learn such skills for the better, beyond simply increasing their likelihood of finding bugs during testing. The decision tree is a much more structured way of expressing a test case's concurrent behaviour than the way 15-410 currently teaches, which is ``think hard until you figure out the buggy interleaving''. Having students use Landslide, even to a small extent, might encourage them to think in this more structured way.

@@

In Section~\ref{sec:related-orthogonal}, we mentioned the technique of symbolic execution for dynamic verification. We believe that symbolic execution could be combined with systematic explorationto help find races in certain obscure code paths that systematic exploration by itself might not even execute.

@@

\begin{figure}[h]
	\small
	\centering
	\begin{tabular}{|l||c|c|}
		\hline
		& \texttt{double\_wait} & \texttt{vanish\_vanish} \\
		\hline\hline
		Exploration time & {\bf 30 minutes} & {\bf 5 minutes} \\
		\hline
		Backtracks & {\bf 1,511} & {\bf 1,401} \\
		\hline
		Decision points & 10,355 & 5,690 \\
		\hline
		Average branch depth & 18.34 & 16.00 \\
		\hline
		Average instructions/branch& 1,975,033 & 1,984,249 \\
		\hline
		Total instructions & 206,757,601 & 24,907,516 \\
		\hline
		Backtrack distance minimum & {\bf 6} & {\bf 2} \\
		\hline
		Backtrack distance maximum & 17 & 16 \\
		\hline
		Backtrack distance average & {\bf 7.38}  & {\bf 3.75} \\
		\hline
		% FIXME: add stddev?
	\end{tabular}
	\caption{Comparison of expensive \texttt{double\_wait} test and cheap \texttt{vanish\_vanish} test. The number of branches and depth of each branch are similar, but \texttt{double\_wait}'s backtracks were significantly longer, causing more total instructions to be executed.}
	\label{fig:nadim-stats}
\end{figure}

>>>>

\begin{figure*}[h]
	\begin{center}
	\includegraphics[width=0.75\textwidth]{wont-modify-vail.jpg}
	\end{center}
	\caption{Former members of Operating Systems course staff lament the difficulty involved in teaching students concurrency debugging skills.}
\end{figure*}

@@

We believe that a tool such as Landslide could change the way students learn such skills for the better, beyond simply increasing their likelihood of finding bugs during testing. The decision tree is a much more structured way of expressing a test case's concurrent behaviour than the way 15-410 currently teaches, which is ``think hard until you figure out the buggy interleaving''. Having students use Landslide, even to a small extent, might encourage them to think in this more structured way.\footnote{
In 15-410, before the kernel project, students also implement a user-space thread library. Just as Landslide could help the students during the kernel project, so too could a user-space systematic exploration tool help during the thread library project.}

@@

In Section~\ref{sec:related-orthogonal}, we mentioned the technique of symbolic execution for dynamic verification. We believe that symbolic execution could be combined with systematic exploration to help find races in certain obscure code paths that systematic exploration by itself might not even execute.

@@

\begin{figure}[h]
	\small
	\centering
	\begin{tabular}{|l||c|c|}
		\hline
		& \texttt{double\_wait} & \texttt{vanish\_vanish} \\
		\hline\hline
		Exploration time & {\bf 30 minutes} & {\bf 5 minutes} \\
		\hline
		Backtracks & {\bf 1,401} & {\bf 1,511} \\
		\hline
		Decision points & 10,355 & 5,690 \\
		\hline
		Average branch depth & 18.34 & 16.00 \\
		\hline
		Average instructions/branch& 1,975,033 & 1,984,249 \\
		\hline
		Total instructions & 206,757,601 & 24,907,516 \\
		\hline
		Backtrack distance minimum & {\bf 6} & {\bf 2} \\
		\hline
		Backtrack distance maximum & 17 & 16 \\
		\hline
		Backtrack distance average & {\bf 7.38}  & {\bf 3.75} \\
		\hline
		% FIXME: add stddev?
	\end{tabular}
	\caption{Comparison of expensive \texttt{double\_wait} test and cheap \texttt{vanish\_vanish} test. The number of branches and depth of each branch are similar, but \texttt{double\_wait}'s backtracks were significantly longer, causing more total instructions to be executed.}
	\label{fig:nadim-stats}
\end{figure}

### 2012-05-12 terminology.tex
continue copyediting

@@

\resume{enumerate}
	\item {\bf Decision point}:
		A point during execution (between two consecutive instructions, precisely) which is deemed ``interesting'' in terms of the likelihood that thread switches at that point will cause concurrency bugs to arise. All thread switches occur at a decision point; some are voluntary reschedules caused by the kernel and others are involuntary preemptions caused by Landslide.
	\item {\bf Transition}:
		A sequence of instructions executed by a single thread between two decision points. At a given decision point, Landslide chooses which thread will run (and causes it to do so), and that thread executes a transition, after which point Landslide will have identified a subsequent decision point
	\item {\bf Decision set} (or {\bf Set of decision points}):
		One or more predicates on the execution state of the guest kernel which identify whether any current state should be a decision point. (For example, when we say ``the decision set including all calls to \texttt{mutex\_lock}'', the corresponding predicate is simply ``did the kernel just invoke \texttt{mutex\_lock}?''.)
	\item {\bf Decision tree} (or {\bf Execution tree}):
		The tree defined by the possibility at each decision point of causing any runnable thread to execute its own transition. Abstractly, this tree comprises all possible states reachable by any interleaving of threads.
		A decision tree is created/discovered during an exploration using a particular decision point set; the set is what the user configures, and the tree is what arises as a result.
	\item {\bf Branch}:
		One particular execution of the test case; a set of decision points with exactly one transition in between each of them, characterising a single interleaving of threads.
	\item {\bf Interleaving}:
		A less precise / more abstract term for a branch, or for a subset of transitions which make up a branch.
	\item {\bf Bug}: An execution state which Landslide identifies as incorrect, which is indicative of illegal behaviour. Each branch of the tree may or may not have a bug in it. See Section~\ref{sec:techniques-bugs}.
	\item {\bf Backtracking}:
		Performed at the end of each branch of the tree, when the test case has finished running. Landslide identifies a decision point from the current branch at which it wants to have chosen a different thread to run, reverts the state of the guest kernel to what it was at that decision point, and causes a different thread to run, hence exploring a different branch/interleaving.
	\item {\bf Decision trace}:
		A list of information about all decision points in a given branch, printed when a bug is found to help the user analyse its cause. Contains, for each decision point, the TID of the thread that was running previously, the TID that was newly chosen, the current instruction pointer, and the stack trace of the previous thread at the point it was switched away from.\footnote{Though Landslide does not provide it, a decision trace may also contain for each transition a list of shared memory accesses that conflicted with other transitions, and/or a listing of happens-before relationships between transitions.}
\end{enumerate}

>>>>

The following terms refer to parts of the guest kernel's execution.

@@

The following terms refer to parts of Landslide's analysis, regardless of the kernel under test's implementation.

\resume{enumerate}
	\item {\bf Decision point}:
		A point during execution (between two consecutive instructions, precisely) which is deemed ``interesting'' in terms of the likelihood that thread switches at that point will cause concurrency bugs to arise. Every thread switch occurs at a decision point; some are voluntary reschedules caused by the kernel and others are involuntary preemptions caused by Landslide.
	\item {\bf Transition}:
		A sequence of instructions executed by a single thread between two decision points. At a given decision point, Landslide chooses which thread will run (and causes it to do so), and that thread executes a transition, after which point Landslide will have identified a subsequent decision point
	\item {\bf Decision set} (or {\bf Set of decision points}):
		One or more predicates on the execution state of the guest kernel which identify whether any current state should be a decision point. (For example, when we say ``the decision set including all calls to \texttt{mutex\_lock}'', the corresponding predicate is simply ``did the kernel just invoke \texttt{mutex\_lock}?''.)
	\item {\bf Decision tree} (or {\bf Execution tree}):
		The tree defined by the possibility at each decision point of causing any runnable thread to execute its own transition. Abstractly, this tree comprises all possible states reachable by any interleaving of threads.
		A decision tree is created/discovered during an exploration using a particular decision point set; the set is what the user configures, and the tree is what arises as a result.
	\item {\bf Branch}:
		One particular execution of the test case; a set of decision points with exactly one transition between each of them, characterising a single interleaving of threads.
	\item {\bf Interleaving}:
		A less precise / more abstract term for a branch, or for a subset of transitions which make up a branch.
	\item {\bf Bug}: An execution state which Landslide identifies as incorrect, which is indicative of illegal behaviour. Each branch of the tree may or may not have a bug in it. See Section~\ref{sec:techniques-bugs}.
	\item {\bf Backtracking}:
		Performed at the end of each branch of the tree, when the test case has finished running. Landslide identifies a decision point from the current branch at which it wants to have chosen a different thread to run, reverts the state of the guest kernel to what it was at that decision point, and causes a different thread to run, hence exploring a different branch/interleaving.
	\item {\bf Decision trace}:
		A list of information about all decision points in a given branch, printed when a bug is found to help the user analyse its cause. Contains, for each decision point, the TID of the thread that was running previously, the TID that was newly chosen, the current instruction pointer, and the stack trace of the previous thread at the point it was switched away from.\footnote{Though Landslide does not provide it, a decision trace may also contain for each transition a list of shared memory accesses that conflicted with other transitions, and/or a listing of happens-before relationships between transitions.}
\end{enumerate}

### 2012-05-12 using.tex
continue copyediting

\subsection{VM}

Landslide does not have much to do with the VM, but the kernel must direct-map most of kernel memory, including the heap, globals, and kernel thread stacks. Landslide needs this to read values out of memory and to identify memory conflicts.\footnote{
It would be simple to support more advanced virtual memory set-ups, in which the same virtual address has different physical memory mappings at different times, but this is not implemented.}

@@

\subsubsection{Important Note about Runqueues}
\label{sec:using-runqueue}
Landslide uses two of these annotations, \texttt{on\_rq()} and \texttt{off\_rq()}, for tracking the state of the scheduler's runqueue at each point during execution.
Note that this refers to the {\em actual runqueue data structure}, which does not necessarily correspond to the abstract ``set of all runnable threads'', depending on whether or not the kernel keeps the current thread on the runqueue or off of it.

>>>>

\subsection{Virtual Memory}

Landslide does not have much to do with the virtual memory system, but the kernel must direct-map most of kernel memory, including the heap, globals, and kernel thread stacks. Landslide needs this to read values out of memory and to identify memory conflicts.\footnote{
It would be simple to support more advanced virtual memory set-ups, in which the same virtual address has different physical memory mappings at different times, but this is not implemented.}

@@

\subsubsection{Important Note about Runqueues}
\label{sec:using-runqueue}
Landslide uses two of these annotations, \texttt{on\_rq()} and \texttt{off\_rq()}, for tracking the state of the guest kernel's scheduler runqueue at each point during execution.
Note that this refers to the {\em actual runqueue data structure}, which does not necessarily correspond to the abstract ``set of all runnable threads'', depending on whether or not the kernel keeps the current thread on the runqueue or off of it.
